Linting Linearity in Core/System FC

### 2022-11-15 proposal.tex
Tweaks
Expand on transformations going wrong

@@

\begin{itemize}
    \item Transformations, occurrences?
    \item Recursive lets
    \item Empty case expression
\end{itemize}

@@

@@

>>>>

\begin{itemize}
    \item Linear types
    \item Linear Haskell
    \item Core
    \item Sistemas de inferência
    \item GADTs e Coercions
\end{itemize}

\section{Related Work}

@@

\begin{itemize}
    \item occurrences? store usage environment in Ids (vars)
    \item Recursive lets (can it be rewritten using fix)
    \item Empty case expression
\end{itemize}

@@

\subsection{Inlining}

If we annotate the let bound variables with their usage and emit that usage when
we come across those variables, we can solve the linearity issues with inlining.

@@

\section{Typing After Transformations}

\subsection{Result of desugar (before optimization)}

This resulting program fails lint with \emph{Linearity failure in lambda: x
'Many $\not\subseteq$ p}
\begin{code}
($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
($!) f !x = f x
~~>
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
            (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of x { __DEFAULT -> f x }
\end{code}

How are strictness annotations typed in the frontend? The issue is this program
consumes to value once to force it, and then again to determine the return
value.

Work better the meaning of consumption (does it mean for a value to be reduced
to WHNF?). Is consuming = forcing? Why is the above program linear?

The resulting program from -ddump-simpl is
\begin{code}
($!)
  :: forall {rep :: GHC.Types.RuntimeRep} a (b :: TYPE rep)
            (p :: GHC.Types.Multiplicity) (q :: GHC.Types.Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: GHC.Types.RuntimeRep))
      (@a)
      (@(b :: TYPE rep_aSJ))
      (@(p :: GHC.Types.Multiplicity))
      (@(q :: GHC.Types.Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of y { __DEFAULT -> f_aDV y }
\end{code}
So in this program we use another name for the case binder, but it still stands
for the resulting of evaluating to WHNF; how is it technically different?

\subsection{In Result of TcGblEnv axioms}

I don't understand this one yet. I would guess it has to do with multiplicity
coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51): [S1
\begin{code}
type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
  Fixup (S1 c f) (S1 _c f) = S1 c f
  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
  Fixup V1 V1 = V1
  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
\end{code}

\subsection{Common Sub-expression Elimination}

Currently, the CSE seems to transform a linear program that pattern matches on
constant and returns the same constant into a program that breaks linearity that
pattern matches on the argument and returns the argument (where in the frontend
a constant equal to the argument would be returned)

Example
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
False && False = False
False && True = False
True && x = x
~~>
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&)
  = \ (x :: Bool) (y :: Bool) ->
      case x of w1 {
        False -> case y of w2 { __DEFAULT -> x };
        True -> y
      }
\end{code}
In the first branch case, we pattern match on y to force it and then return $x$
rather than the constant $False$.

At first look, the resulting program is impossible to typecheck.

Idea: if $x$ was anotated with some information regarding the CSE then perhaps
it could be typechecked (in a system that considered said anotations)

Confusingly: the resulting optimised program from running the complete simplifier
and outputting with -ddump-simpl actually does what we expected it to regarding
the constants and linearity. So is the issue from running linear lint after a
particular CSE but it would be fine in the end?
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&)
  = \ (x :: Bool) (y :: Bool) ->
      case x of {
        False -> case y of { __DEFAULT -> GHC.Types.False };
        True -> y
      }
\end{code}
